Constructive forcing

### 2023-09-17 slides-herrsching2023.tex
More notes

    Recall that a given ring~$R$ or group can be extended in various ways, to include
    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
    similar such extensions, but not of rings but of universes (traditionally set-sized models
    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
    of intuitionistic set theories, or models of type theories, or even models
    of arithmetic).

@@

    Instead, we should embrace the multiverse of all models of set theory and
    explore how the truth value of statements of interest change when we
    travel the multiverse (for instance, by passing from some universe to one
    of its forcing extensions).

@@

  \jnote{3}{
    The real numbers don't contain a number~$i$ such that~$i^2 = 0$. Still, for
    many results in real analysis, it is convenient to broaden our notion of
    existence and pass to the complex numbers; the imaginary unit is a
    \emph{mathematical phantom} in the
    \fixedhref{http://www.wra1th.plus.com/gcw/math/MathPhant.html}{sense of
    Gavin Wraith}, a useful tool helping us deduce results about real numbers.
    Nowadays there are few ontological concerns about the imaginary unit: We
    understand that, in the end, every statement about complex numbers can be
    recast as a statement about pairs of real numbers.

@@

    In the context of constructive mathematics, however, the generic sequence
    turns out to be quite useful. It can be used to cast in a familiar naive
    language---the language of infinite sequences---definitions, results and
    proofs which use constructively more appropriate inductively defined
    notions. We will discuss an example below.
  }

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals}
  \only<1-7>{\textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

@@

  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
    \node (0) at (0,1) {$(0) = \{0\}$};
    \node (1) at (0,5) {$(1) = \ZZ$};
    \node (2) at (-2,4) {$(2)$};
    \node [right of=2] (3) {$(3)$};
    \node [below of=2] (4) {$(4)$};
    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
    \node [right of=3] (5) {$(5)$};
    \node [right of=5] (7) {$(7)$};
    \node [right of=7] (7d) {$\ldots$\phantom{(}};
    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
    \mmm)$ \\ \medskip \textbullet \quad $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
    \node [below of=4] (8) {$(8)$};
    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
    \draw (0) -- (8);
    \draw (0) -- (8d);
    \draw (0) -- (6);
    \draw (2) -- (1);
    \draw (3) -- (1);
    \draw (5) -- (1);
    \draw (7) -- (1);
    \draw (7d) -- (1);
    \draw (4) -- (2);
    \draw (8) -- (4);
    \draw (6) -- (2);
    \draw (6) -- (3);
    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple!30]
        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}\par}\par}

>>>>

    Recall that a given ring~$R$ or group can be extended in various ways, to include
    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
    similar such extensions, but not of rings but of universes (traditionally set-sized models
    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models, or models
    of intuitionistic set theories, or models of type theories, or even models
    of arithmetic).

@@

    Instead, we should embrace the multiverse of all models of set theory and
    explore how the truth values of statements of interest change when we
    travel the multiverse (for instance, by passing from a universe to one
    of its forcing extensions).

@@

  \jnote{3}{
    The real numbers don't contain a number~$i$ such that~$i^2 = 0$. Still, for
    many results in real analysis, it is convenient to broaden our notion of
    existence and pass to the complex plane; the imaginary unit is a
    \emph{mathematical phantom} in the
    \fixedhref{http://www.wra1th.plus.com/gcw/math/MathPhant.html}{sense of
    Gavin Wraith}, a useful tool helping us deduce results about real numbers.
    Nowadays there are few ontological concerns about the imaginary unit: We
    understand that, in the end, every statement about complex numbers can be
    recast as a statement about pairs of real numbers.

@@

    In the context of constructive mathematics, however, the generic sequence
    turns out to be quite useful. It can be used to cast in a familiar naive
    language---the language of infinite sequences---definitions, results and
    proofs from constructive combinatorics which use constructively more
    appropriate inductively defined notions. We will discuss an example below.
  }

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals}
  \jnote{1-}{
    The theorem on the slide is a generalization of a fact from undergraduate
    linear algebra: Over a field, no surjective matrix can have more rows than
    columns. (``Surjective'' here means that the induced linear map is
    surjective.)
  }

  \jnote{6-}{
    The slide presents a standard proof as offered by most textbooks on
    commutative algebra. The proof is quite efficient from a viewpoint of
    mathematical organization, as it quickly succeeds in reducing to the field
    situation. As such, it is short and memorable.

    However, the proof can also be critized for appealing to the transfinite
    two times; the methods of the proof are at odds with the concreteness of
    the statement of the theorem---from given equations witnessing
    surjectivity,~$Mv_i = e_i$, we are asked to deduce the equation~$1 = 0$.

    For this reason, the theorem and its classical proof are often used as case
    studies for tools and techniques aiming to extract constructive content
    from classical proofs. One such technique employs constructive forcing.
    The first(?) constructive proof, found directly without using extraction
    techniques, is laid out in a
    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
  }

  \only<1-7>{\textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

@@

  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
    \node (0) at (0,1) {$(0) = \{0\}$};
    \node (1) at (0,5) {$(1) = \ZZ$};
    \node (2) at (-2,4) {$(2)$};
    \node [right of=2] (3) {$(3)$};
    \node [below of=2] (4) {$(4)$};
    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
    \node [right of=3] (5) {$(5)$};
    \node [right of=5] (7) {$(7)$};
    \node [right of=7] (7d) {$\ldots$\phantom{(}};
    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
    \node [below of=4] (8) {$(8)$};
    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
    \draw (0) -- (8);
    \draw (0) -- (8d);
    \draw (0) -- (6);
    \draw (2) -- (1);
    \draw (3) -- (1);
    \draw (5) -- (1);
    \draw (7) -- (1);
    \draw (7d) -- (1);
    \draw (4) -- (2);
    \draw (8) -- (4);
    \draw (6) -- (2);
    \draw (6) -- (3);
    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple!30]
        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}\par}\par}
